---
title: "Palmdale School District 19-20 and 20-21 Data"
author: "Jessica Rico, Gurshan Rai, and Chinyere Ogugu"
date: "`r Sys.Date()`"
output: 
  html_notebook: 
    toc: yes
  html_document: 
    keep_md: yes
    toc: yes
runtime: shiny
---
Our group project subject was on sociocultural issues, and we chose to analyze the effects of the switch to distance learning due to the pandemic on middle school students. Schools had to abruptly shut down and switch to remote learning, leaving a lot of students behind due to inaccessibility to internet or computer access and the lack of in class engagement with teachers and other students . We decided to study the impact of school closures on student academic achievement. We were lucky that one of our group members works for a middle school and she was able to provide us with data for the statistics for attendance and GPA before and after the pandemic.

# Our Questions about Palmdale School District

How did the switch to distance learning due to the pandemic affect attendance?

How did the switch to distance learning affect grade distributions?

What conclusions can we draw from data? What kind of future policy changes, ramifications with returning to in person classes, and issues with inequity can be drawn from our analysis?

# Our Data

We used publicly available datasets published by the Palmdale School District. All data was obtained with permission of school administrators and all Personally Identifiable Information was already omitted from our datasets.  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
if (!require("tidyverse")) install.packages('tidyverse')
```


```{r}
library(tidyverse)
library(rmarkdown)
library(shiny)
library(shinydashboard)
library(skimr)
library(janitor)
library(here)
library(RColorBrewer)
library(paletteer)
library(ggthemes)
library(naniar)
library(dplyr)
library(ggpubr)
library(ggplot2)
library(here)
options(scipen=999)
```



# Attendance Rates by Month Data Loading and Manipulation
```{r}
attrates<- readr::read_csv("data/att_rates.csv")
attrates<- janitor::clean_names(attrates)
attrates$year<- as.factor(attrates$year)
attrates$month<- as.factor(attrates$month)
attrates
```
### attrates
This data set shows the percent of student attendance for attendance months 1-5 by year. This also shows the number of students that both enrolled and withdrew from the school site in each of these months along with the total number of enrollment at each of these months. (Note: the months are attendance months rather than calendar months)

```{r}
attrates_long <- attrates %>% 
  pivot_longer(
    3:6,
    names_to="attendance_statistic",
    values_to="value"
  )
attrates_long
#write.csv(attrates_long, "data/attrateslong.csv")
```

# Attendance Codes by Year Loading and Manipulation
```{r}
attreason <- readr::read_csv("data/att_reason_reports.csv")
attreason
```

```{r}
attcodes<- attreason%>%
  pivot_longer(-year, names_to = "att_code", values_to = "occurence")
attcodes$year<- as.factor(attcodes$year)
attcodes
```
### attcodes
This dataset shows the usage of attendance codes by year. These attendance codes are used when we enter the reason behind an absence. Some, such as DR, ILL, are excused while others like UNV, and VUN are not excused. These factor in when it comes to determining our attendance percentages. The codes ILL19, AVP, NWFA, DAC, and NCA were introduced for virtual/distance learning due to the pandemic. 

# Attendance Codes by Month & Year Loading and Manipulation
```{r}
codesmonth<- readr::read_csv("data/attendance codes by month.csv")
codesmonth<- janitor::clean_names(codesmonth)
codesmonth$month <- as.factor(codesmonth$month)
codesmonth$year <- as.factor(codesmonth$year)
codesmonth
```
### codesmonth
This dataset shows the usage of attendance codes by month and year rather than showing the total counts for each code by year (as seen above in "attcodes"). 

```{r}
codesmonth_wide <- codesmonth %>% 
  pivot_wider(
    names_from = "code",
    values_from = "used"
  )
codesmonth_wide
```

```{r}
codes_summary <- codesmonth%>%
  group_by(code, year)%>%
  summarise(total=sum(used))
```

# Attendance Numbers Shiny App 

```{r}
attrates_long <- readr::read_csv("attrateslong.csv")

ui <- dashboardPage(skin ="black",
    dashboardHeader(title = "Attendance Data"),
    dashboardSidebar(disable = T),
    dashboardBody(
        fluidRow(
            box(title = "Plot Options", width = 3,
                selectInput("x", "Select Term", choices = c("attendance_rate_percent", "student_add", "student_drop"),
                            selected = "attendance_rate_percent"),
            ), # close the first box
            box(title = "Attendance Rates, Enrollment and Withdrawal Data by Month", width=6,
                plotOutput("plot", width = "500px", height = "300px")
            ) # close the second box
        ) # close the row
    ) # close the dashboard body
) # close the ui
server <- function(input, output, session) { 
    output$plot <- renderPlot({
        attrates_long %>% 
            filter(attendance_statistic==input$x) %>% 
            ggplot(aes(x=month, y=value)) +
            geom_col(color="black", fill="darkcyan", alpha=.5)+
            facet_wrap(~year)+
            #geom_col(color="black", fill="slateblue2", alpha=1)+
            theme_gdocs()+
            labs(x="Month",y="Count")
        
    })
    
    # stop the app when we close it
    session$onSessionEnded(stopApp)
}
shinyApp(ui, server)
```
This shinyApp shows the attendance rates, enrollment and withdrawals by month and year. At first sight, there does not appear to be many differences between the two years. We decided to do a deeper analysis into the attendance rates that are used to determine the actual attendance rates. 


# Attendance Code Usage by Month and Year Plots

```{r, message=FALSE, echo=FALSE, cache = TRUE}
codesmonth%>%
  ggplot(aes(x=code, y=used, fill=code))+
  geom_col()+
  facet_grid(year~month)+
  theme_light()+
  theme(axis.text.x = element_text(angle = 90, hjust=1, size = 8))+
  labs(title = "Number of Times Attendance Codes Used by Month and Year",
       x = "Attendance Code",
       y= "Number of Times Used")
```

Here we can see a comparison of the number of times that each attendance code was used by month and year. The high number of AVP's immediately stands out. These are the main type of unexcused absence for the year 2020. 
```{r}
codessubset<- subset(codesmonth, code==c("AVP","UNV"))
```

```{r}
p2<- codessubset%>%
  ggplot(aes(x=code, y=used, fill=year))+
  geom_col(position="dodge")

```

```{r}
p2+theme_pander()+
  theme(axis.text.x = element_text(angle = 0, hjust=1))+
  labs(title = "Attendance Codes Used in First Semester by Year",
       x = "Attendance Code",
       y= "Number of Times Used",
       fill= "Year")
```
AVP stands for "absent virtual participation"; this essentially means that the student was absent from their class period for an unknown reason (no parent contact, Dr note, etc). This code was introduced in 2020 due to the shift to distance learning. In regular year, an unknown reason for an absence is documented using the code UNV which stands for unverified. This plot is showing the dramatic increase in total number of unknown absences for the first semester of 2020 vs the total number for the first semester of 2019. With all of these absences, we were curious to see what the impact was on academic performance.

# First Semester GPA by Grade Level and Year App

```{r}
gpa_2_long <- readr::read_csv("gpa_2_long.csv")

ui <- dashboardPage(skin ="black",
    dashboardHeader(title = "GPA Shiny App"),
    dashboardSidebar(disable = T),
    dashboardBody(
        fluidRow(
            box(title = "Plot Options", width = 2,
                radioButtons("x", "Select Grade", choices = c("6", "7", "8"), 
                             selected = "6"),
                radioButtons("y", "Select Term", choices = c("1", "2"),
                             selected = "1"),
                radioButtons("z", "Select Year", choices = c("2019", "2020"),
                             selected = "2019"),
            ), # close the first box
            box(title = "GPA Data From Years 19-20 and 20-21", width=6,
                plotOutput("plot", width = "500px", height = "300px")
            ) # close the second box
        ) # close the row
    ) # close the dashboard body
) # close the ui

server <- function(input, output, session) { 
    
    output$plot <- renderPlot({
        gpa_2_long %>% 
            filter(grade==input$x & term==input$y & year==input$z) %>% 
            ggplot(aes(x=gpa, fill=year)) +
            geom_density(color="black", fill="darkcyan", alpha=.5)+
            theme_gdocs()+
            labs(x="GPA", y="Student Count")
        
    })
    
    # stop the app when we close it
    session$onSessionEnded(stopApp)
}

shinyApp(ui, server)
```
All the grades listed are from the first semester of both years. Term 1 represents progress report grades and Term 2 is the final semester grade received. This shinyApp shows the trends for each grade level by term and year. 

# In Conclusion



